---
title: "Example of GAM spline fit with simultaneous CI"
output:
  html_notebook:
    highlight: haddock
    theme: default
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---


# Preamble
```{r preamble}
library(here)
here()

# load packages
library(tidyverse)
library(mgcv)

# bright color blind palette:  https://personal.sron.nl/~pault/ 
cblack <- "#000004FF"
cblue <- "#3366AA"
cteal <- "#11AA99"
cgreen <- "#66AA55"
cchartr <- "#CCCC55"
cmagent <- "#992288"
cred <- "#EE3333"
corange <- "#EEA722"
cyellow <- "#FFEE33"
cgrey <- "#777777"

```


# Load a dataset

This is publicly available antibody data from Cambodia based on this article:
https://wwwnc.cdc.gov/eid/article/24/7/17-1928-f1

Dataset is available here:
https://osf.io/2kr8b/

```{r load data}
#-------------------------------
# load the cambodia serology data
#-------------------------------
d <- readRDS(here("data","cambodia_serology_public.rds"))

#-------------------------------
# select tetanus response
# and convert to log10 MFI
# format cluster ID to factor
# create a dummy variable equal
# to 1 for all obs as a trick to
# get marginal predictions by
# age while still including REs
# for cluster
#-------------------------------
d <- d %>%
  select(psuid,age,ttmb) %>%
  mutate(psuid=factor(psuid),
         logmfi=log10(ifelse(ttmb<0,1,ttmb)),
         dummy=1)

```


# Function for a simultaneous CI around a spline curve
```{r simultaneous CI}
#----------------------------------
# simulataneous CIs for GAMs
# estimated by resampling the 
# Baysian posterior estimates of
# the variance-covariance matrix
# assuming that it is multivariate normal
# the function below also estimates 
# the unconditional variance-covariance
# matrix, Vb=vcov(x,unconditional=TRUE), 
# which allows for undertainty in the actual
# estimated mean as well 
# (Marra & Wood 2012 Scandinavian Journal of Statistics, 
#  Vol. 39: 53–74, 2012, doi: 10.1111/j.1467-9469.2011.00760.x )
# simultaneous CIs provide much better coverage than pointwise CIs
# see: http://www.fromthebottomoftheheap.net/2016/12/15/simultaneous-interval-revisited/
#----------------------------------

gamCI <- function(m,newdata,nreps=10000) {
  require(mgcv)
  require(dplyr)
  Vb <- vcov(m,unconditional = TRUE)
  pred <- predict(m, newdata, se.fit = TRUE)
  fit <- pred$fit
  se.fit <- pred$se.fit
  BUdiff <- MASS::mvrnorm(n=nreps, mu = rep(0, nrow(Vb)), Sigma = Vb)
  Cg <- predict(m, newdata, type = "lpmatrix")
  simDev <- Cg %*% t(BUdiff)
  absDev <- abs(sweep(simDev, 1, se.fit, FUN = "/"))
  masd <- apply(absDev, 2L, max)
  crit <- quantile(masd, prob = 0.95, type = 8)
  pred <- data.frame(newdata,fit=pred$fit,se.fit=pred$se.fit)
  pred <- mutate(pred,
                 uprP = fit + (2 * se.fit),
                 lwrP = fit - (2 * se.fit),
                 uprS = fit + (crit * se.fit),
                 lwrS = fit - (crit * se.fit)
  )
  return(pred)
}

```

# Estimate mean antibody level by age

```{r agecurve}
#----------------------------------
# ignore clustering (i.e., no random effect for cluster)
#----------------------------------
# fit GAM with a spline for age
fit1 <- mgcv::gam(logmfi~s(age, bs="cr"),data=d)
# estimate simultaneous CIs around the curve
fit1ci <- gamCI(m=fit1,newdata=d,nreps=10000)


#----------------------------------
# add a random effect for clustering
#----------------------------------
# fit GAM with a spline for age
fit2 <- mgcv::gam(logmfi~s(age, bs="cr")+s(psuid,bs="re",by=dummy),data=d)
# estimate simultaneous CIs around the curve
# for the prediction data, set the dummy to 0 to 
# zero out all of the random effects
# see posts on Stack Exchange for explanation:
# https://stats.stackexchange.com/questions/131106/predicting-with-random-effects-in-mgcv-gam/131116#131116
# https://stats.stackexchange.com/questions/189384/predicting-mean-smooth-in-gam-with-smooth-by-random-factor-interaction
newd <- d %>% mutate(dummy=0)
fit2ci <- gamCI(m=fit2,newdata=newd,nreps=10000)

```

# Plot results

The simultaneous CIs increase the width of the band.

In this particular example, accounting for cluster has almost no effect on the width of the simulaneous CIs.  Possibly because there are 100 clusters, so sample size is still very large.

```{r figures, fig.width=3, fig.height=2}

p1 <- ggplot(data=fit1ci,aes(x=age)) +
  # geom_point(aes(y=logmfi),alpha=0.2,color=cblack) +
  geom_line(aes(y=fit),color=cblack) +
  geom_ribbon(aes(ymin=lwrS,ymax=uprS),alpha=0.2,color=NA,fill=cred) +
  geom_ribbon(aes(ymin=lwrP,ymax=uprP),alpha=0.3,color=NA,fill=cred) +
  labs(title="Tetanus antibody response by age, ignoring clustering") +
  theme_minimal()
p1

p2 <- ggplot(data=fit2ci,aes(x=age)) +
  # geom_point(aes(y=logmfi),alpha=0.2,color=cblack) +
  geom_line(aes(y=fit),color=cblack) +
  geom_ribbon(aes(ymin=lwrS,ymax=uprS),alpha=0.2,color=NA,fill=cteal) +
  geom_ribbon(aes(ymin=lwrP,ymax=uprP),alpha=0.3,color=NA,fill=cteal) +
  labs(title="Tetanus antibody response by age, accounting for clustering") +
  theme_minimal()
p2

p3 <- ggplot(data=fit1ci,aes(x=age)) +
  geom_line(aes(y=fit),color=cblack) +
  geom_ribbon(aes(ymin=lwrP,ymax=uprP),alpha=0.2,color=cgrey,fill=cteal) +
  geom_ribbon(data=fit2ci,aes(ymin=lwrP,ymax=uprP),alpha=0.3,color=cgrey,lty="dashed",fill=cgreen) +
  labs(title="Tetanus antibody response by age, comparison of pointwise CIs") +
  theme_minimal()
p3

p4 <- ggplot(data=fit1ci,aes(x=age)) +
  geom_line(aes(y=fit),color=cblack) +
  geom_ribbon(aes(ymin=lwrS,ymax=uprS),alpha=0.2,color=cgrey,fill=cteal) +
  geom_ribbon(data=fit2ci,aes(ymin=lwrS,ymax=uprS),alpha=0.3,color=cgrey,lty="dashed",fill=cgreen) +
  labs(title="Tetanus antibody response by age, comparison of simultaneous CIs") +
  theme_minimal()
p4
```



